\documentclass[letterpaper,12pt,leqno]{article}
\usepackage{paper,math,notes}
\available{https://pascalmichaillat.org/x/}
\hypersetup{pdftitle={Problem Set on Mathematics for Macroeconomics}}

\begin{document}

\title{Problem Set on Mathematics for Macroeconomics}
\author{Pascal Michaillat}
\date{}

\begin{titlepage}
\maketitle
\end{titlepage}

\section*{Problem 1}

Let $\a \in (0,1)$, $\d  \in (0,1)$, $\rho \in (0,1)$, and $\s>0$. Impose that $\rho+\d<1$. Given $k(0)$, we want to find the function $c(t) $ to maximize 
\begin{equation*}
\int_{0}^{+\infty }e^{-\rho\cdot  t}\cdot \frac{c(t)^{1-\s}-1}{1-\s} dt,
\end{equation*}
subject to the law of motion
\begin{equation*}
\dot{k}(t) =k(t)^{\a}-c(t)-\d \cdot k(t).
\end{equation*}

\begin{enumerate}

\item  Which variable do you choose as a state variable? Which variable do you choose as a control variable? Write down the current-value Hamiltonian and derive the optimality conditions.

\item The Euler equation is the first-order differential equation that characterizes the optimal function $c(t)$. Determine the Euler equation.

\item  Suppose $\a =1$ and $\s =1$. Show that the system describing the optimal functions $\{k(t),c(t)\}$ reduces to a linear, homogenous system of first-order differential equations. Show that the system is unstable by computing the eigenvalues. 

\item  Suppose $\a <1$ and $\s >0$. Show that the system describing the optimal functions $\{k(t),c(t)\}$ reduces to a nonlinear system of first-order differential equations. Use a phase-diagram to show that the steady state of the system is a saddle point. Explain how you draw the phase diagram.

\end{enumerate}

\section*{Problem 2}

Let $\b \in (0,1)$ and $r>0$. Given $k_{0}>0$, we want to find a collection of sequences $\{c_{t},k_{t+1}\}_{t=0}^{+\infty}$ to maximize 
\begin{equation*}
\sum_{t=0}^{\infty }\b^{t} \cdot \ln(c_{t}), 
\end{equation*}
subject to the constraints
\begin{equation*}
k_{t+1}=(1+r)\cdot k_{t}-c_{t}
\end{equation*}
for all $t\geq 0$.

\paragraph{Lagragian} We first solve the maximization problem using the Lagrangian method.

\begin{enumerate}
\item Write down the Lagrangian of the problem.
\item Derive the first-order condition(s) of the maximization problem.
\item Derive the Euler equation.
\end{enumerate}

\paragraph{Dynamic Programming} Next we solve the maximization problem using the dynamic programming method.

\begin{enumerate}\setcounter{enumi}{3}

\item Which variable do you choose as a state variable? Which variable do you choose as a control variable? Write down the Bellman equation.

\item  Derive the first-order condition associated with the Bellman equation.

\item  Derive the Benveniste-Scheinkman equation.

\item  Derive the Euler equation. Compare it with the Euler equation obtained with the Lagrangian method and discuss.

\item Suppose that the policy function takes the form $h(k)=A\cdot (1+r)\cdot k$ where $A\in (0,1) $. Derive $A$.

\item  Suppose that the value function takes the form $V(k)=B+D\cdot \ln(k),$ where $B$ and $D$ are constants. Using the expression for the policy function that you derived in the previous question, derive $B$ and $D$.

\end{enumerate}
\end{document}